Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-17 cordillera-cycle.tex
Add minor corrections from Martin in Sect. 1--3.

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000})
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2014}.

Because climate conditions over the last glacial cycle are subject to
considerable uncertainty, our palaeoclimate forcing is a simplistic approach
including temperature and precipitation fields derived from a
present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
Seguinot.etal.2014} supplemented by a lapse-rate correction
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotopes records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotopes
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific ocean
sediment cores \citep{Herbert.etal.2001}. Model output is compared to
geomorphological evidence in terms of the timing and extent of glaciation and
patterns of deglaciation.

@@

Each simulation starts from assumed ice-free conditions at 120\,000 years ago
(120\,ka), and runs to present. Our modelling domain of 1\,500 by 3\,000\,km
encompasses the entire area covered by the Cordilleran ice sheet at the LGM
(Fig.~\ref{fig:locmap}). The simulations were run on two distinct grids, using
a lower horizontal resolution of 10\,km, and a higher horizontal resolution of
6\,km. These computations were performed on 16 to 128 computing cores at the
Swedish National Supercomputing Centre.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
coast \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice limits at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
In the rest of this paper, we occasionally refer to different model
runs by the name of the proxy record used for the palaeo-temperature forcing.

@@

thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
    Small change in notation to match with the kappa},
on the one hand, and
for differences between the the ice flow model basal topography, $b$, and the
NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.

@@

Despite large differences in the input atmospheric forcing (temperature offset
time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
consistently attained between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent atmospheric forcing
(Table~\ref{tab:extrema})\renote{
    Hopefully this new table will assess several of your comments at once.}.

Simulations forced by the Greenland ice core palaeo-temperature\renote{
    From here on, I prepend ``palaeo-temperature'' to ``record(s)'' in place of
    ``climate'' as you (Arjen) originally suggested. If we are to precise what
    type of ``records'' we use, let's be clear about it.}
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in smaller\renote{
    Until we run out of other ajectives, I don't think that ``modest'' and
    ``generous'' are adequate terms for describing ice volumes.}
modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
More broadly speaking, they produce lower ice volume variability
throughout the simulation length. The NGRIP atmospheric forcing is the only one
that results in a larger ice volume during MIS~4 than during the last glacial
maximum (MIS~2).

While simulations driven by GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
informed by NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
slower deglaciation than modelled using other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records produce a
rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also yields a rapid
deglaciation but in three steps, interrupted by two stand-stills and readvances
of the ice margin.

@@

Despite such differences in the timing of attained volume maxima and mimima,
snapshots of model output show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1\,500\,km-long central divide
located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
Mountains. The volumes of these modelled last glacial maximum ice sheets are
contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
(s.l.e; Table~\ref{tab:extrema}).
A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.

>>>>

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000})
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

Because climate conditions over the last glacial cycle are subject to
considerable uncertainty, our palaeoclimate forcing is a simplistic approach
including temperature and precipitation fields derived from a
present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
Seguinot.etal.2014} supplemented by a lapse-rate correction
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotopes records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotopes
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific ocean
sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
output to geological evidence and discuss the timing and extent of glaciation
and the patterns of deglaciation.

@@

Each simulation starts from assumed ice-free conditions at 120\,000 years ago
(120\,ka), and runs to the present. Our modelling domain of 1\,500 by
3\,000\,km
encompasses the entire area covered by the Cordilleran ice sheet at the LGM
(Fig.~\ref{fig:locmap}). The simulations were run on two distinct grids, using
a lower horizontal resolution of 10\,km, and a higher horizontal resolution of
6\,km. These computations were performed on 16 to 128 computing cores at the
Swedish National Supercomputing Centre.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
\renote{
    ``Monthly climatology'' is a common term in meteorology to design a set
    of variables aggregated over several decades, and assumed to represent
    some climate state.},
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice limits at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
    Small change in notation to match with the kappa},
on the one hand, and for differences between the basal topography of the ice
flow model, $b$, and the
NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
In the rest of this paper, we refer to different model runs by the name of the
proxy record used for the palaeo-temperature forcing.

@@

Despite large differences in the input atmospheric forcing (temperature offset
time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent atmospheric forcing
(Table~\ref{tab:extrema})\renote{
    Hopefully this new table will assess several of your comments at once.}.

Simulations forced by the Greenland ice core palaeo-temperature\renote{
    From here on, I prepend ``palaeo-temperature'' to ``record(s)'' in place of
    ``climate'' as you (Arjen) originally suggested. If we are to precise what
    type of ``records'' we use, let's be clear about it.}
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in smaller\renote{
    Until we run out of other ajectives, I don't think that ``modest'' and
    ``generous'' are adequate terms for describing ice volumes.}
modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
More broadly speaking, they produce lower ice volume variability
throughout the simulation length. The NGRIP atmospheric forcing is the only one
that results in a larger ice volume during MIS~4 than during the LGM (MIS~2).

While simulations driven by GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
informed by NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records results in a
rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also result in a rapid
deglaciation but in three steps, interrupted by two stand-stills and readvances
of the ice margin.

@@

Despite such differences in the timing of attained volume maxima and mimima,
all the model runs show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1\,500\,km-long central divide
located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
Mountains. The volumes of these modelled last glacial maximum ice sheets are
contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
(s.l.e; Table~\ref{tab:extrema}).
A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.
